Unifying Lower Bounds on Prediction Dimension of Consistent Convex Surrogates

### 2021-01-26 colt-21/cvx-flats.tex
more in 6

\newcommand{\ellbar}{\underline{\ell}}
\newcommand{\lbar}{\underline{L}} % couldn't do L* while proofreading...
\newcommand{\iden}{\mathrm{iden}}
\newcommand{\im}{\mathrm{im}}
\newcommand{\Var}{\mathrm{Var}}

@@

\paragraph{Example: Variance.}
As a warm-up, let us see how to show $\eliccvx(\Var)=2$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2~\citep{osband1985providing,lambert2018elicitation}.
It is interesting to note that, to the best of our knowledge, even this simple bound is novel.
While intuitively obvious, the lower bound of 2 is not trivial.
In particular, the well-known fact that the variance is not elicitable does not yield this lower bound, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet[Remark 1]{frongillo2018elicitation}.

Let $\Y\subseteq\reals$.
Observe that the variance of $Y$ is the Bayes risk of squared loss $L(r,y) = (r-y)^2$; as $L$ elicits the mean $\Gamma(p) = \E_p [Y]$, we have $\Var_p[Y] = \{\E_p (\E_pY - Y)^2 \} = \min_{r\in\reals} \E_p L(r,Y)$.
To apply Theorem~\ref{thm:bayes-risk-lower-bound}, we take this $L$ and $\Gamma$ and $d=1$, and must choose an appropriate level set $\Gamma_r$
\raf{Fill in details: $\P$ has two distributions with different means, and two with the same mean but different variances}

\raf{The bound is tight}
We may elicit the first two moments via the convex $L(r,y) = (r_1-y)^2 + (r_2-y^2)^2$, and recover the variance via $\psi(r) = r_2-r_1^2$.

\raf{Summarize in a Cor?}

@@

\paragraph{Example: Entropy and Norms.}
\jessie{Move to Appendix?}
For another application, let us see how Theorem~\ref{thm:bayes-risk-lower-bound} implies $\eliccvx(G) = n-1$ for any strictly concave function $G:\simplex\to\reals$ of the distribution, including most entropy functions.
In other words, there is no convex loss function allowing one to consistently estimate conditional entropy using fewer dimensions than required to estimate the conditional distribution itself.
This observation also extends to norms; given any $k>1$, $G(p) = \|p\|_k^k$ is strictly convex, and hence $\eliccvx(\|\cdot\|_k) = n-1$, as otherwise we could link to $G$ via $\psi(r) = r^k$.
These results illustrate the power of our technique.

To show the bound, recall that $G$ is the Bayes risk of a proper loss defined by $L(p,y) = G(p) + s_p \cdot (\delta_y - p)$, where $\delta_y(y') = \ones\{y=y'\}$ is the point distribution on $y$ and $-s_p \in \partial (-G)(p)$ is a subgradient of $-G$ at $p$~\citep{gneiting2007strictly,reid2009surrogate,frongillo2014general}.
Intuitively, since $L$ elicits the identity property $\Gamma(p)=p$, Theorem~\ref{thm:bayes-risk-lower-bound} should therefore give us a maximal lower bound ($n-1$), as the level sets of are $\Gamma$ have codimension $n-1$.
For technical reasons, however, we need to drop a dimension from the prediction space, e.g.\ via the bijection $\varphi(p) = (p_1,\ldots,p_{n-1})$, defining $L'(p',y) = L(\varphi^{-1}(p'),y)$ which elicits $\Gamma'(p) = \varphi(p)$ but still has Bayes risk $G$.
Now the conditions of Theorem~\ref{thm:bayes-risk-lower-bound} are easily checked, where again $p\in\relint\simplex$ is the uniform distribution and $r=\varphi(p)$: (i) is trivial, (ii) $\codim(\affhull(\Gamma_r')) = \codim(\{p\}) = n-1$, and (iii) $\Gamma_r'$ is a singleton.
As with the variance example, the lower bound is easily matched, e.g.\ by $L_2(p',y) = \|\varphi^{-1}(p')-\delta_y\|_2^2$, which is convex in $p'$, and the link $\psi(p') = G(\varphi^{-1}(p'))$.

>>>>

\newcommand{\ellbar}{\underline{\ell}}
\newcommand{\lbar}{\underline{L}} % couldn't do L* while proofreading...
\newcommand{\iden}{\mathrm{iden}}
\newcommand{\im}{\mathrm{im}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\CVaR}{\mathrm{CVaR}}

@@

We now illustrate the theorem with two important examples: variance and conditional value at risk.
Several other applications from \citet{frongillo2018elicitation}, such as spectral risk measures, entropy, and norms, follow similarly.

\paragraph{Example: Variance.}
As a warm-up, let us see how to show $\eliccvx(\Var)=2$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2.%~\citep{osband1985providing,lambert2018elicitation}.
This lower bound follows from Theorem~\ref{thm:bayes-risk-lower-bound} as the variance is the Bayes risk of squared loss $L(r,y) = (r-y)^2$, which elicits the mean $\Gamma(p) = \E_p Y$.
Interestingly, while intuitively obvious, even this simple result is novel.
In particular, the well-known fact that the variance is not elicitable does not yield a lower bound of 2, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet[Remark 1]{frongillo2018elicitation}.

\begin{corollary}
  \label{cor:variance}
  Let $\P\subseteq\simplex$ for $\Y\subseteq \reals$, such that there exist $p,q,q'\in\P$ with $\E_p Y = \E_q Y \neq \E_{q'} Y$ and $\Var(p) \neq \Var(q)$.
  \raft{Note to self: I think this condition is tight actually: if there is no such triple in $\P$, I think $\eliccvx(\Var) \leq 1$ (i.e., $\Var$ is constant or a function of the mean)}
  Then $\eliccvx(\Var)=2$.
\end{corollary}
\begin{proof}
  For the upper bound, we may elicit the first two moments via the convex loss $L(r,y) = (r_1-y)^2 + (r_2-y^2)^2$, and recover the variance via $\psi(r) = r_2-r_1^2$, giving $\eliccvx(\Var) \leq 2$.
  Now for the lower bound.
  Without loss of generality, $\E_qY < \E_{q'}Y$.
  Let $r = \tfrac 1 2 \E_qY + \tfrac 1 2 \E_{q'}Y$, and define $V:\Y\to\reals, y\mapsto y-r$.
  Then $\zeros{V} = \{p'\in\P \mid \E_{p'}Y=r\} = \Gamma_r$ where $\Gamma:p'\mapsto \E_{p'}Y$ is the mean.
  As $\E_q Y < r < \E_{q'} Y$, we conclude $\E_q V < 0 < \E_{q'} V$.
  We have now satisfied Condition~\ref{cond:v-interior} for $d=1$.
  To apply Theorem~\ref{thm:bayes-risk-lower-bound}, it remains to show that $\Var$ is non-constant on $\Gamma_r$.
  % and the identity $\Var(p') = \E_{p'} Y^2 - (\E_{p'} Y)^2$
  By our assumptions and the definition of $\Var$, we have $\E_p Y^2 \neq \E_q Y^2$.
  Letting $p_1 = \tfrac 1 2 q + \tfrac 1 2 q'$, $p_2 = \tfrac 1 2 p + \tfrac 1 2 q'$, we have $\E_{p_i}Y = r$ for $i\in\{1,2\}$, but $\E_{p_1} Y^2 = \tfrac 1 2 \E_qY^2 + \tfrac 1 2 \E_{q'}Y^2 \neq \tfrac 1 2 \E_pY^2 + \tfrac 1 2 \E_{q'}Y^2 = \E_{p_2}Y^2$.
  As $p_1,p_2$ have the same mean but different second moments, we conclude $\Var(p_1) \neq \Var(p_2)$.
\end{proof}

@@

\citet{frongillo2018elicitation} observe that one of the most prominent financial risk measures, the conditional value at risk (CVaR), can be expressed as a Bayes risk.
In particular, for $0 < \alpha < 1$, we may define
\begin{align}
  \CVaR_\alpha(p)
  &= \inf_{r\in\reals} \E_p\left\{ \tfrac 1 \alpha(r-Y)\ones_{r\geq Y}- r\right\}~,
  \label{eq:elic-complex-es-3}
\end{align}
which is the Bayes risk of the transformed pinball loss $L_\alpha(r,y) = \tfrac 1 \alpha(r-y)\ones_{r\geq y}-r$.
In turn, $L_\alpha$ elicits the $\alpha$-quantile $q_\alpha(p) = \raf{DEF}$.

\begin{corollary}
  \label{cor:spectral-risks}
  Let $\P$ \raf{four distributions: $p_0,p_1,p_2$ w/ increasing quantiles and $p_1'$ with the same quantile as $p_1$ but different tail}.
  Then $\eliccvx(\CVaR_\alpha) = 2$.
\end{corollary}
